The persuasion of symbolic elements in government public communications, and their moderation by personal relevance

### 2018-07-05 paper3_relevance and symbols_01.Rmd
Add files via upload

I further explored the robustness of these findings additional analyses, reported in the supplementary appendix. First, in addition to the above operationalizations of personal relevance, I also relied on subjects' subjective reporting. Second, I further explored the effect of personal relevance on subjects inclination to scrutinize the policy plans, using additional measures for message elaboration available in the survey: the time participants spent on scanning the policy plans and their score in a memory test. I also tested whether personal relevance decreased the delta (within-subjects) between trust in strong and weak policy plans presented to each subject. Third, I added a set of individual-level controls, and factors related to personal relevance of air-pollution, and to personal interest in environmental issues. Finally, I replicated my analyses on the unfiltered sample. None of these multiple robust analyses supports my theoretical hypotheses.

@@

Extant Public-Administration research has demonstrated that government communication can affect citizens' trust not only by conveying substantive information, but also by using familiar symbolic elements. This study contributes to this literature by exploring the postulation, derived from social psychology, that people are less susceptible to symbols and more likely to scrutinize the substantive information in the communication the more they perceive it as personally relevant. I put this theory to rigorous empirical test in the context of government communications, using a randomized survey experiment focusing on the Israeli Environmental Protection Ministry. I further focused on its policy plan regarding air pollution in a specific area, to capture natural variation in citizens' perceptions of the personal relevance. In addition to utilizing these natural differences, I also used manipulated the personal relevance of the policy plans.

Whilst the effect of perceived personal relevance on message processing has been substantiated by experiment-based studies in social psychology and marketing, I did not find support for this hypothesis the context of citizens' responses to government communications in this study. Subjects residing in the polluted Haifa-Bay indeed reported that they perceive the policy plan for reducing air-pollution in that area as highly relevant for their lives, compared to those who reside in the center. However, despite these differences, Haifa-bay subjects did not pay more attention to the logical quality of the policy plan, and they were affected by the symbols in the communication not less than those from the center. If anything, the results suggest that the symbols had a greater effect on Haifa-bay subjects. Similarly, priming subjects to think about the personal relevance of environmetal policy, and of the consequences of air-pollution in particular, did not moderate the effects of the symbols and information in the communications. Even those ,  The comparison between those who were assigned to the treatment

How can we reconcile these unexpected findings with the proposed theory? Given that the moderating role of personal relevance on peripheral cues was substantiated by numerous studies in social psychology, I would not argue that we should abandon this postulation based on this one study. One plausible explanation is that although Haifa-bay residents perceived the air-pollution policy as more relevant, their motivation to elaborate was not high enough to cause them to avoid peripheral cues. Certainly, the air-pollution in their area concerns them a great deal, as they point out themselves. However this issue represents a *continuous* concern that they may have got used to over the time, as opposed a new issue that emerged recently. If this explanation holds, this entails that citizens may resist the emotional effect of symbols only in rare circumstances, which are characterized by very high levels of personal relevance. These latter findings are also consistent with the findings of @alonbarkat_2017. In that study, the scholars assessed citizens' trust in the government electricity company, and its shaping by its familiar symbol, given citizens' recent personal experience of prolonged blackouts. Whilst their study was not purposely designed to test the interaction between symbols and personal relevance, one can assume that those who personally experienced blackouts perceived electricity service as more personally relevant. Still, they found that citizens who experienced blackouts were influenced by the symbol, as much as those who did not.

>>>>

I further explored the robustness of these findings through additional analyses, reported in the supplementary appendix. First, in addition to the above operationalizations of personal relevance, I also relied on subjects' subjective reporting. Second, I further explored the effect of personal relevance on subjects inclination to scrutinize the policy plans, using additional measures for message elaboration available in the survey: the time participants spent on scanning the policy plans and their score in a memory test. I also tested whether personal relevance decreased the delta (within-subjects) between trust in strong and weak policy plans presented to each subject. Third, I added a set of individual-level controls, and factors related to personal relevance of air-pollution, and to personal interest in environmental issues. Finally, I replicated my analyses on the unfiltered sample. None of these multiple robust analyses supports my theoretical hypotheses.

@@

Extant Public-Administration research has demonstrated that government communication can affect citizens' trust not only by conveying substantive information, but also by using familiar symbolic elements. This study contributes to this literature by exploring the postulation, derived from social psychology, that people are less susceptible to symbols and more likely to scrutinize the substantive information in the communication the more they perceive it as personally relevant. I put this theory to rigorous empirical test in the context of government communications, using a randomized survey experiment focusing on the Israeli Environmental Protection Ministry. I further focused on its policy plan regarding air pollution in a specific area, to capture natural variation in citizens' perceptions of the personal relevance. In addition to utilizing these natural differences, I also used manipulated the personal relevance of the policy plans. To the best of my knowledge, this is the first study to assess the  moderating role of perceived personal relevance on message elaboration and the effect of symbol in the administrative context.

Whilst the effect of perceived personal relevance on message processing has been substantiated by experiment-based studies in social psychology and marketing, I did not find support for this hypothesis the context of citizens' responses to government communications in this study. Subjects residing in the polluted Haifa-Bay indeed reported that they perceive the policy plan for reducing air-pollution in that area as highly relevant for their lives, compared to those who reside in the center. However, despite these differences, Haifa-bay subjects did not pay more attention to the logical quality of the policy plan, and they were affected by the symbols in the communication not less than those from the center. If anything, the results suggest that the symbols had a greater effect on Haifa-bay subjects. Priming subjects to think about the personal relevance of environmental policy, and of the consequences of air-pollution in particular, also did not moderate the effects of the symbols and information in the communications.

How can we explain the finding that the positive effect of the real symbols was stronger among Haifa-Bay residents? The fact that this pattern evident in both policy plans suggests that these differences are not explained by perceived personal relevance. Rather, these differences could be explained, by the lower baseline of Haifa-Bay subjects' trust in both policies, presumably due to their negative prior attitudes towards EPM. Accordingly, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa-Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects). Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust. This tentative explanation is supported by the fact that the interaction remains intact when controlling for demographic differences between the areas.[^footnote_ceiling_effect]

[^footnote_ceiling_effect]: This tentative explanation is also supported by the comparison of the percentages of subjects who reported very low levels of trust in the policies. This comparison shows a pattern of decreased marginal reduction. In the Haifa-Bay, the percentage of those who reported a trust score of less than 2 in the air-pollution policy decreased by from `r ((filter(SVIVA2_01_haifa,SYMBOL==0,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==0)%>%nrow())*100)%>%round(1)`% in the no symbols group to `r ((filter(SVIVA2_01_haifa,SYMBOL==2,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==2)%>%nrow())*100)%>%round(1)`%. In the center, only `r ((filter(SVIVA2_01_center,SYMBOL==0,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==0)%>%nrow())*100)%>%round(1)`% of the no symbols group reported a trust score lower than 2 in the first place, and this percentage dropped to `r ((filter(SVIVA2_01_center,SYMBOL==2,TRUST_air_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==2)%>%nrow())*100)%>%round(1)`% among the real symbols group. With regard to the recycling policy, the percentage decreased from `r ((filter(SVIVA2_01_haifa,SYMBOL==0,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==0)%>%nrow())*100)%>%round(1)`% to `r ((filter(SVIVA2_01_haifa,SYMBOL==2,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_haifa,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the Haifa-bay and from `r ((filter(SVIVA2_01_center,SYMBOL==0,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==0)%>%nrow())*100)%>%round(1)`% to `r ((filter(SVIVA2_01_center,SYMBOL==2,TRUST_waste_INDEX<2) %>% nrow())/(filter(SVIVA2_01_center,SYMBOL==2)%>%nrow())*100)%>%round(1)`% in the center.

How can we reconcile the findings of this study with the proposed theory? One plausible explanation is that despite the relatively high involvement of  Haifa-bay residents, they were still not motivated enough to use the central route and to disregard the peripheral cues. If this explanation holds, it entails that citizens may resist the emotional effect of symbols only in rare circumstances, which are characterized by extremely high levels of personal relevance. Indeed, Petty and Cacioppo have admitted that the high and low personal relevance conditions generated in their experimental settings were deliberately created for theory testing purposes, and can be rarely found in the world outside the laboratory [@petty_1986, p. 206]. The findings in this study are also consistent with the findings of @alonbarkat_2017, where they found that citizens who recently personally experienced prolonged blackouts (and presumably perceived electricity services as personally relevant) were nonetheless significantly affected by a symbol of the government-owned electricity company.

The results of this study, if generalizable, have significant normative implications. They challenge the optimistic view of government direct communications with citizens as a means to inform citizens about government organizations actions and promote transparency and accountability. Instead, they suggest that more often than not, citizens rely on symbols and other peripheral cues in public communications, including when these communications concern issues that have direct implications on citizens' personal welfare. This inclination is in tension with the assumption that citizens should form their opinions about their government and about public affairs, based on rational thinking [@dahl_1973; @soroka_2010]. It also exacerbates the concern that government organizations will strategically use symbolic communications to manipulate public opinion, and cause citizens to view them more favorably regardless (or even in contrast) to their actual capacities and achievements.

Notwithstanding these conclusions, I acknowledge the limitations of this study, which focuses, empirically, on one specific case. Future studies may put these findings to additional test by examining additional cases of public organizations and policy areas, as well as different forms of communications. Additionally, the comparison between the two areas is limited insofar as the results suggest that they have diverging prior beliefs about the subject. A comparison between two groups that have similar prior beliefs would be more reliable. Finally, as noted above, one can question the validity of relevance manipulation, which has not yielded significant effect on subjects' subjective reporting on the perceived personal relevance of the communications. Still, the main results are robust to different model specifications, and operationalizations of personal relevance and message elaboration, which substantiate their validity.
